---
title: "Data science in the Tidyverse"
author: "[Insert your name]"
output:
  html_document:
    toc: true
---

*This document is part of the workshop **Introduction to R & Data** by Utrecht University RDM Support. *

# Introduction

<!-- About the data, about the purpose -->

## Exercises

Each of the following themes will have two kinds of exercises:

- Exercises labeled with an M are mandatory; without it you will not be able to proceed to the next theme. Make sure you complete it before moving on.
- Exercises labeled with an O are challenges for you to explore the material further. While they are recommended, they are not necessary for you to finish before moving on to the next step.


## Technical requirements

We are going to use a series of packages that are part of the Tidyverse. These packages were designed with data science in mind. We can load them all at once using the following code:

```{r} 
library(tidyverse) 
```

# Import 

## Exercise (M)

In this exercise, we read the dataset `HUJI_Crane_Israel_GPRS.csv` into the memory of the computer. 

You will need to know three things to complete this exercise:
- The **function** to use to read data into R;
- The **location** of your data;
- The **delimiter** of the data.

You can find the dataset in the `data` folder of your project. Note that the file location (argument `file`) should be `data/HUJI_Crane_Israel_GPRS.csv`, as you are referring to its location (or 'path') relative to the location of this script.

The delimiter is the character that separates the values in a tabular data file. There are various ways to find out how the datasets is delimited. One way from inside Rstudio is to use the 'Import Dataset' button (in your 'Environment') window, choose 'from Text (readr)', locate your data, and play around with the settings. An **example** of the code you can use should now appear in the bottom right window. (This will also include a function!)

Complete the code in the chunk below:

```{r}
# load data
penguins <- read_delim("data/penguins_raw.tsv", delim = "\t")
```



## Exercise (O)

<!-- Load additional data from an excel file. Maybe an excel file with information on the islands? -->

`read_csv` and `read_delim` are useful functions for most tabular data files, but they are unable to read excel files. For this, we need an additional package.

The package `readxl` is a tidyverse package to read Excel files. This package
does not load by default when you call `library(tidyverse)`. 

Load this package with the function `library()`:

Included in the *package* `readxl` (a package is a collection of functions) is the *function* `read_excel`. You can learn how to use this function by exploring its help function. To do this, type `?read_excel` in your console.

Use the function `read_excel` to read the additional observations in data file `crane_additional_observations.xlsx` (an Excel file!), also located in your `data` folder, into R.

Complete the code in the chunk below:

```{r}


```


The structure of the dataset (i.e. columns, data types, number of observations) is very important for your initial data exploration. Tidyverse has the function `glimpse()` to output the structure of a dataframe or tibble. 

Output the structure of both datasets in the cell below. What do `<chr>`, `<dbl>` and `<int>` mean? Are these values what you expect?

## Optional exercise (+++) - Parse datetime columns

Date and time variables can be very challenging to work with in programming languages (including R). In the `data_crane` dataset, there is a variable `eobs_start_timestamp`. This column is not recognized as a date variable, as you can see when you glimpse the data: 

```{r}
glimpse(data_crane)
```

The following code shows how you can columns with a specific data type. As in excercise 1.I (a), you can use the 'Import Dataset' button (in your 'Environment') window to play around with the settings, and see an example of the code this corresponds to in the bottom right window.

Insert the correct format for the date column:

```{r}

read_csv('data/HUJI_Crane_Israel_GPRS.csv',
  col_types = cols(
    . = col_guess(),
    eobs_start_timestamp = col_date(format="## insert correct format here ##")
  ))
  
```



# Tidy

### Exercise (M)

Make a selection of all observations of crane 'L6037' where the variable `eobs_status` is not missing. To do this, you can use the function `is.na()`, but you will have to invert its results. Search in your help window for the operator `!`, to get an idea on how to do this.

How many records satisfy this condition?

```{r}

# filter data
penguins <- filter(penguins, !is.na(Sex))

```


## Exercise (O)

practice Mutate

```{r}
penguins <- penguins %>% 
  mutate(corrected_flipper = `Flipper Length (mm)`/`Body Mass (g)`)
```


## Exercise (M)
<!-- Select the right columns

https://tidyselect.r-lib.org/reference/language.html -->

```{r}
penguins <- select(penguins, Species, Sex, `Individual ID`, starts_with("Culmen"))
```

## Optional exercise (++) - Join datasets

In this exercise we will append the crane dataset with the additional measures. We will do a full join, combining all data from `data_crane` with all data from `data_crane_additional`. (Note though that both datasets have the same unique identifiers, so both a right join and a left join would achieve the same results.)

Join the two data frames by adding to the code below. What column should you use as join key (the `by=` argument)?

*Tip: for more information about joining data, take a look at the cheat sheet element 'Combine Tables', or the recommended reading for this chapter.*

```{r}

data_crane_with_measures <- full_join(## enter your code here ##

```


## Exercise (O)
<!-- Explain -->

```{r}
penguins <- rename(penguins,
                   length = `Culmen Length (mm)`,
                   depth = `Culmen Depth (mm)`)


```



## Exercise (O)
<!-- Explain -->

```{r}


penguins <- pivot_longer(penguins, 
                         cols = c(length, depth),
                         names_to = "culmen_element", values_to = "measurement_mm")

```



### Exercise 2 (O)
<!-- Explain -->

```{r}
# optional: pivot back to original



```



### Exercise 2 (O)
<!-- Explain -->

```{r}
penguins_summary <- penguins %>%
  group_by(Species, culmen_element) %>%
  summarize(average_mm = mean(measurement_mm),
            sd = sd(measurement_mm))

```



### Exercise 2 (O)
<!-- Explain -->


Save the tibble `data_crane` to a `.csv` file. Use `;` as delimiter. To do this, you can use the function `write_delim`. Again, you can make use of its help function if you want to understand how the function works: type `?write_delim` in your console.

*Tip: Use a different file name, otherwise you overwrite the original data*
*Tip: You can save the file to a specific folder by including it in the path, just like you did when you read the file in exercise I-1. For instance:* `data/other_file_name.csv` *will save your file in the* `data` *folder*.
```{r}
# optional: write back to file

```



### Exercise 2 (O)
<!-- Explain -->

```{r}
ggplot(penguins_summary,
       aes(x = culmen_element,
           y = average_mm, 
           fill = Species)) +
  geom_bar(position=position_dodge(), stat="identity") +
  geom_errorbar(aes(ymin=average_mm-sd, 
                      ymax=average_mm+sd), 
                  width=.1, position=position_dodge(.9)) +
  labs(title = "Penguin culmen measurements",
       x = "Measurement",
       y = "Size (mm)") +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  theme_bw()

```



### Exercise 2 (O)
<!-- Explain -->

```{r}
penguins_clean <- palmerpenguins::penguins %>%
  filter(!is.na(sex))

```

The power of ggplot is in the use of different layers that build up an image. Your plot consists of your data, its mapping to a plot, and one or multiple geometric layers that translate the data to a visual representation.

Play with that translation in showing the cranes' ground speed by the angle in which they are heading. Try various geometric layers and see what they do. Examples are `geom_point`, `geom_line`, `geom_count` and `geom_density_2d`, but feel free to search for others!

*Tip: write `geom_` in the search field of the help tab in RStudio to find more geometric layers.*

Complete the following code. (Note that the plot is first generated as p, but not printed yet: it is saved as a variable. You can then always call p and add layers to it.)

```{r}

ggplot(penguins_clean, aes(x = bill_length_mm,
                           y = bill_depth_mm,
                           color = species)) +
  geom_point() +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  stat_density2d() +
  theme_bw()
```



```{r}



ggplot(penguins_clean, aes(x = body_mass_g,
                           y = flipper_length_mm,
                           color = species)) +
  geom_point() +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  facet_grid(sex ~ island) +
  theme_bw()



```







