# Tibbles and data import

This part of the workshop covers chapter 10, 11.1, 11.2, 11.4, 11.6 of *R for Data Science (Wickham & Gorlemunt, 2017)*. 

## Introduction

In this section, you'll learn how to read plain-text rectangular files into R. Here, we'll only scratch the surface of data import, but many of the principles will translate to other forms of data. We'll finish with a few pointers to packages that are useful for other types of data.

Working with data provided by R packages is a great way to learn the tools of data science, but at some point you want to stop learning and start working with your own data. In this chapter, you'll learn how to read plain-text rectangular files into R. Here, we'll only scratch the surface of data import, but many of the principles will translate to other forms of data. We'll finish with a few pointers to packages that are useful for other types of data.

### Prerequisites

In this chapter, you'll learn how to load flat files in R with the __readr__ package, which is part of the core tidyverse.

```{r}
library(tidyverse)
```

Import readr as a standalone package with the following command:
```{r}
library(readr)
```

## Getting started

Most of readr's functions are concerned with turning flat files into data frames:

* `read_csv()` reads comma delimited files, `read_csv2()` reads semicolon
  separated files (common in countries where `,` is used as the decimal place),
  `read_tsv()` reads tab delimited files, and `read_delim()` reads in files
  with any delimiter.

* `read_fwf()` reads fixed width files. You can specify fields either by their
  widths with `fwf_widths()` or their position with `fwf_positions()`.
  `read_table()` reads a common variation of fixed width files where columns
  are separated by white space.

These functions all have similar syntax: once you've mastered one, you can use the others with ease. For the rest of this chapter we'll focus on `read_csv()`. Not only are csv files one of the most common forms of data storage, but once you understand `read_csv()`, you can easily apply your knowledge to all the other functions in readr.

The first argument to `read_csv()` is the most important: it's the path to the file to read.

```{r, message = TRUE}
data <- read_csv("data/Huji_JNF_Crane_israel_GPRS.csv")
```

When you run `read_csv()` it prints out a column specification that gives the name and type of each column. That's an important part of readr, which we'll come back to in [parsing a file].

1.  The data might not have column names. You can use `col_names = FALSE` to
    tell `read_csv()` not to treat the first row as headings, and instead
    label them sequentially from `X1` to `Xn`:
    
    ```{r}
    read_csv("1,2,3\n4,5,6", col_names = FALSE)
    ```
    
    (`"\n"` is a convenient shortcut for adding a new line. You'll learn more
    about it and other types of string escape in [string basics].)
    
    Alternatively you can pass `col_names` a character vector which will be
    used as the column names:
    
    ```{r}
    read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
    ```

Another option that commonly needs tweaking is `na`: this specifies the value (or values) that are used to represent missing values in your file:

```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```

This is all you need to know to read ~75% of CSV files that you'll encounter in practice. You can also easily adapt what you've learned to read tab separated files with `read_tsv()` and fixed width files with `read_fwf()`. To read in more challenging files, you'll need to learn more about how readr parses each column, turning them into R vectors.

### Compared to base R

If you've used R before, you might wonder why we're not using `read.csv()`. There are a few good reasons to favour readr functions over the base equivalents:

>  * They are typically much faster (~10x) than their base equivalents.
>    Long running jobs have a progress bar, so you can see what's happening. 
>    If you're looking for raw speed, try `data.table::fread()`. It doesn't fit 
>    quite so well into the tidyverse, but it can be quite a bit faster.
>  
>  * They produce tibbles, they don't convert character vectors to factors,
>    use row names, or munge the column names. These are common sources of
>    frustration with the base R functions.
>  
>  * They are more reproducible. Base R functions inherit some behaviour from
>    your operating system and environment variables, so import code that works 
>    on your computer might not work on someone else's.

### Exercises

1.  What function would you use to read a file where fields were separated with  
    "|"?
    
1.  Apart from `file`, `skip`, and `comment`, what other arguments do
    `read_csv()` and `read_tsv()` have in common?
    
1.  What are the most important arguments to `read_fwf()`?
   
1.  Sometimes strings in a CSV file contain commas. To prevent them from
    causing problems they need to be surrounded by a quoting character, like
    `"` or `'`. By convention, `read_csv()` assumes that the quoting
    character will be `"`, and if you want to change it you'll need to
    use `read_delim()` instead. What arguments do you need to specify
    to read the following text into a data frame?
    
    ```{r, eval = FALSE}
    "x,y\n1,'a,b'"
    ```


## [Parsing a vector](http://r4ds.had.co.nz/data-import.html#parsing-a-vector)

If there are many parsing failures, you'll need to use `problems()` to get the complete set. This returns a tibble, which you can then manipulate with dplyr.

```{r}
# problems(x)
```


You pick between three parsers depending on whether you want a date (the number of days since 1970-01-01), a date-time (the number of seconds since midnight 1970-01-01), or a time (the number of seconds since midnight). When called without any additional arguments:

*   `parse_datetime()` expects an ISO8601 date-time. ISO8601 is an
    international standard in which the components of a date are
    organised from biggest to smallest: year, month, day, hour, minute, 
    second.
    
    ```{r}
    parse_datetime("2010-10-01T2010")
    # If time is omitted, it will be set to midnight
    parse_datetime("20101010")
    ```
    
    This is the most important date/time standard, and if you work with
    dates and times frequently, I recommend reading
    <https://en.wikipedia.org/wiki/ISO_8601>
    
*   `parse_date()` expects a four digit year, a `-` or `/`, the month, a `-` 
    or `/`, then the day:
    
    ```{r}
    parse_date("2010-10-01")
    ```

*   `parse_time()` expects the hour, `:`, minutes, optionally `:` and seconds, 
    and an optional am/pm specifier:
  
    ```{r}
    library(hms)
    parse_time("01:10 am")
    parse_time("20:10:01")
    ```
    
    Base R doesn't have a great built in class for time data, so we use 
    the one provided in the hms package.

If these defaults don't work for your data you can supply your own date-time `format`, built up of the following pieces:

Year
: `%Y` (4 digits). 
: `%y` (2 digits); 00-69 -> 2000-2069, 70-99 -> 1970-1999.

Month
: `%m` (2 digits).
: `%b` (abbreviated name, like "Jan").
: `%B` (full name, "January").

Day
: `%d` (2 digits).
: `%e` (optional leading space).

Time
: `%H` 0-23 hour.
: `%I` 0-12, must be used with `%p`.
: `%p` AM/PM indicator.
: `%M` minutes.
: `%S` integer seconds.
: `%OS` real seconds. 
: `%Z` Time zone (as name, e.g. `America/Chicago`). Beware of abbreviations:
  if you're American, note that "EST" is a Canadian time zone that does not
  have daylight savings time. It is _not_ Eastern Standard Time! We'll
  come back to this [time zones].
: `%z` (as offset from UTC, e.g. `+0800`). 

Non-digits
: `%.` skips one non-digit character.
: `%*` skips any number of non-digits.

The best way to figure out the correct format is to create a few examples in a character vector, and test with one of the parsing functions. For example:

```{r}
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")
```

If you're using `%b` or `%B` with non-English month names, you'll need to set the  `lang` argument to `locale()`. See the list of built-in languages in `date_names_langs()`, or if your language is not already included, create your own with `date_names()`.

```{r}
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```

### Exercises

1.  What are the most important arguments to `locale()`? 

1.  What's the difference between `read_csv()` and `read_csv2()`?






## Parsing a file

Now that you've learned how to parse an individual vector, it's time to return to the beginning and explore how readr parses a file. There are two new things that you'll learn about in this section:

1. How readr automatically guesses the type of each column.
1. How to override the default specification.

### Strategy

readr uses a heuristic to figure out the type of each column: it reads the first 1000 rows and uses some (moderately conservative) heuristics to figure out the type of each column. You can emulate this process with a character vector using `guess_parser()`, which returns readr's best guess, and `parse_guess()` which uses that guess to parse the column:

```{r}
guess_parser("2010-10-01")
guess_parser("15:01")
guess_parser(c("TRUE", "FALSE"))
guess_parser(c("1", "5", "9"))
guess_parser(c("12,352,561"))

str(parse_guess("2010-10-10"))
```

The heuristic tries each of the following types, stopping when it finds a match:

* logical: contains only "F", "T", "FALSE", or "TRUE".
* integer: contains only numeric characters (and `-`).
* double: contains only valid doubles (including numbers like `4.5e-5`).
* number: contains valid doubles with the grouping mark inside.
* time: matches the default `time_format`.
* date: matches the default `date_format`.
* date-time: any ISO8601 date.

If none of these rules apply, then the column will stay as a vector of strings.

### Problems

These defaults don't always work for larger files. There are two basic problems:

1.  The first thousand rows might be a special case, and readr guesses
    a type that is not sufficiently general. For example, you might have 
    a column of doubles that only contains integers in the first 1000 rows. 

1.  The column might contain a lot of missing values. If the first 1000
    rows contain only `NA`s, readr will guess that it's a character 
    vector, whereas you probably want to parse it as something more
    specific.

There are two printed outputs: the column specification generated by looking at the first 1000 rows, and the first five parsing failures. It's always a good idea to explicitly pull out the `problems()`, so you can explore them in more depth:

```{r}
# problems(challenge)
```

A good strategy is to work column by column until there are no problems remaining. Here we can see that there are a lot of parsing problems with the `x` column - there are trailing characters after the integer value. That suggests we need to use a double parser instead.

To fix the call, start by copying and pasting the column specification into your original call:

```{r, eval = FALSE}
data <- read_csv(
  "data/Huji_JNF_Crane_israel_GPRS.csv", 
  col_types = cols(
    `tag-local-identifier` = col_character(),
    visible = col_logical()
  )
)

spec(data)
```

Then you can tweak the type of the `x` column:

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_character()
  )
)
```

That fixes the first problem, but if we look at the last few rows, you'll see that they're dates stored in a character vector:

```{r}
tail(challenge)
```

You can fix that by specifying that `y` is a date column:

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge)
```

Hadley Wickham:

> I highly recommend always supplying `col_types`, building up from the print-out provided by readr. This ensures that you have a consistent and reproducible data import script. If you rely on the default guesses and your data changes, readr will continue to read it in. If you want to be really strict, use `stop_for_problems()`: that will throw an error and stop your script if there are any parsing problems.




### Other strategies

There are a few other general strategies to help you parse files:

*   In the previous example, we just got unlucky: if we look at just
    one more row than the default, we can correctly parse in one shot:
   
    ```{r}
    challenge2 <- read_csv(readr_example("challenge.csv"), guess_max = 1001)
    challenge2
    ```

*   Sometimes it's easier to diagnose problems if you just read in all
    the columns as character vectors:
   
    ```{r}
    challenge2 <- read_csv(readr_example("challenge.csv"), 
      col_types = cols(.default = col_character())
    )
    ```
    
    This is particularly useful in conjunction with `type_convert()`,
    which applies the parsing heuristics to the character columns in a data
    frame.

    ```{r}
    df <- tribble(
      ~x,  ~y,
      "1", "1.21",
      "2", "2.32",
      "3", "4.56"
    )
    df
    
    # Note the column types
    type_convert(df)
    ```
    
*   If you're reading a very large file, you might want to set `n_max` to
    a smallish number like 10,000 or 100,000. That will accelerate your 
    iterations while you eliminate common problems.

*   If you're having major parsing problems, sometimes it's easier
    to just read into a character vector of lines with `read_lines()`,
    or even a character vector of length 1 with `read_file()`. Then you
    can use the string parsing skills you'll learn later to parse
    more exotic formats.

## Writing to a file

readr also comes with two useful functions for writing data back to disk: `write_csv()` and `write_tsv()`. Both functions increase the chances of the output file being read back in correctly by:

* Always encoding strings in UTF-8.
  
* Saving dates and date-times in ISO8601 format so they are easily
  parsed elsewhere.

If you want to export a csv file to Excel, use `write_excel_csv()` --- this writes a special character (a "byte order mark") at the start of the file which tells Excel that you're using the UTF-8 encoding.

The most important arguments are `x` (the data frame to save), and `path` (the location to save it). You can also specify how missing values are written with `na`, and if you want to `append` to an existing file.

```{r, eval = FALSE}
write_csv(challenge, "challenge.csv")
```

Note that the type information is lost when you save to csv:

```{r, warning = FALSE}
challenge
write_csv(challenge, "challenge-2.csv")
read_csv("challenge-2.csv")
```


## Other types of data

To get other types of data into R, we recommend starting with the tidyverse packages listed below. They're certainly not perfect, but they are a good place to start. For rectangular data:

* __haven__ reads SPSS, Stata, and SAS files.

* __readxl__ reads excel files (both `.xls` and `.xlsx`).

* __DBI__, along with a database specific backend (e.g. __RMySQL__, 
  __RSQLite__, __RPostgreSQL__ etc) allows you to run SQL queries against a 
  database and return a data frame.

For hierarchical data: use __jsonlite__ (by Jeroen Ooms) for json, and __xml2__ for XML. Jenny Bryan has some excellent worked examples at <https://jennybc.github.io/purrr-tutorial/>.

For other file types, try the [R data import/export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html) and the [__rio__](https://github.com/leeper/rio) package.
