---
title: "Modern R with tidyverse [Solutions]"
author: "Jonathan de Bruin & Barbara Vreede"
output:
  pdf_document:
    toc: yes
    latex_engine: xelatex
  html_document:
    toc: yes
---

*This document is part of the workshop **Introduction to R & Data** by Utrecht University RDM Support. *

```{r} 
library(tidyverse) 
```

# 1. Read and save data


### Basic exercise I - Read data into R

#### a) Find the delimiter of the dataset `Huji_JNF_Crane_israel_GPRS.csv`

Use the data import tool of RStudio or use a text editor to inspect the document. The delimiter is `,`.

#### b) Read the data `Huji_JNF_Crane_israel_GPRS.csv` into R

```{r error=TRUE}

data_crane <- read_csv('data/Huji_JNF_Crane_israel_GPRS.csv')

head(data_crane)

```

#### c) Load `readxl` (to read Excel files)

```{r error=TRUE}

library(readxl)

# open the help function
?read_excel

```

#### c) Read the additional observations `Huji_JNF_additional_observations.xlsx`.

```{r error=TRUE}

data_crane_additional <- read_excel('data/Huji_JNF_additional_observations.xlsx')

head(data_crane_additional)
```


### Basic exercise II - Dataset properties 

```{r error=TRUE}

glimpse(data_crane)
glimpse(data_crane_additional)

```

### Reading exercise - `readr` versus base R

### Optional exercise (+) - Save data to a CSV file with delimiter `;`. 

```{r error=TRUE}
# create a directory for the output file
if (!dir.exists('tmp')){
  dir.create("tmp")
}

write_delim(data_crane, 'tmp/data_crane_csv_file.csv', delim = ';')

```

### Optional exercise (++) - Read SPSS, SAS, Excel and STATA data files

```{r}
library("haven") # to read and write SPSS, STATA and SAS files
```

#### a) Write tibble `data_crane` to SPSS, SAS, STATA data files. 

```{r}
# create a directory
if (!dir.exists('tmp')){
  dir.create("tmp")
}
# read and write files
write_sav(data_crane, file.path("tmp", "crane_spss.sav"))
read_sav(file.path("tmp", "crane_spss.sav"))

# write_sas(data_crane, file.path("tmp", "crane_sas.sas7bdat"))
# read_sas(file.path("tmp", "crane_sas.sas7bdat"))

# write_dta(data_crane, file.path("tmp", "crane_stata.dta"))
# read_dta(file.path("tmp", "crane_stata.dta"))

```



#### b) Write tibble `data_crane` to Excel. 

This is not possible with `tidyverse` at the moment. `readxl` only support Excel file reading. 
This is not a problem for a researcher, because we don't use Excel, do we?

### Optional exercise (+++) - Parse datetime columns

No solutions available at the moment.

# 2. Data visualisation

### Basic exercise I - Quick plots of the `data_crane`.

#### a) Single column plots

Make a quick plot of at least 5 columns in the `data_crane` dataset.
```{r error=TRUE}

qplot(timestamp, data=data_crane)
qplot(mag_magnetic_field_raw_x, data=data_crane)
qplot(mag_magnetic_field_raw_y, data=data_crane)
qplot(tag_voltage, data=data_crane)
qplot(eobs_status, data=data_crane)


```
#### b) Two column in plots

```{r error=TRUE}

qplot(heading, acceleration_raw_x, data=data_crane)
qplot(acceleration_raw_x, acceleration_raw_y, data=data_crane)
qplot(acceleration_raw_x, acceleration_raw_z, data=data_crane)
qplot(location_long, location_lat, data=data_crane)



```


### Basic exercise II - Using ggplot for plotting

```{r error=TRUE}

ggplot(data_crane, aes(heading, ground_speed)) + 
  geom_count()

```

### Reading exercise - Statistical layers for graphs.

Statistical layers reveal a strong power of ggplot. The following graph illustrates this:

```{r echo=FALSE, message=FALSE, width=4, height=3}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() + geom_smooth()
```

Read [Chapter 3.6](https://r4ds.had.co.nz/data-visualisation.html#geometric-objects)  of [R for Data
Science](http://r4ds.had.co.nz/) and experiment with the function `geom_smooth()` and `stat_smooth()`. What is the difference between both functions? 


### Optional exercise (+) - Scale axes

Scaling of the axes with a ggplot is easy. Take a look at the *data-visualization* cheat sheet. In the bottom right corner of the cheat sheet, you will find the code needed to scale the axes. 

Zoom in on the acceleration between -1000 and 1000 for both the x-axis and y-axis. 

```{r error=TRUE}
ggplot(data_crane, aes(acceleration_raw_x, acceleration_raw_y)) + 
  geom_point()+ 
  scale_x_continuous(limits = c(-1000, 1000)) +
  scale_y_continuous(limits = c(-1000, 1000))
```



### Optional exercise (++) - Plot the crane positions on a map

#### a) Install package the `maps`

```{r error=TRUE}

# install.packages('maps')
library(maps)

```

#### b) Plot the crane data on a map.

```{r error=TRUE}
world_map_polygon <- map_data("world2")

ggplot(data_crane) +
  geom_map(data = world_map_polygon, map= world_map_polygon, aes(x=long, y = lat, map_id = region)) + 
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(25, 60)) + 
  geom_point(data = data_crane, aes(x = location_long, y = location_lat))

```


#### c) Use an individual identifier to colour the different cranes.

```{r error=TRUE}

world_map_polygon <- map_data("world2")

ggplot(data_crane) +
  geom_map(data = world_map_polygon, map= world_map_polygon, aes(long, lat, map_id = region)) + 
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(25, 60)) + 
  geom_point(data = data_crane, aes(x = location_long, y = location_lat, colour=individual_local_identifier))


```

### Optional exercise (+++) - Create facets. 

```{r error=TRUE, fig.height=15, fig.width=6}

ggplot(data_crane, aes(mag_magnetic_field_raw_x, mag_magnetic_field_raw_y)) +
  geom_point() + 
  facet_wrap(~individual_local_identifier)

```


# 3. Data transformation

### Basic exercise I - Subset data (single crane)

#### a) Make a selection of all observations of crane 'L6037'.

```{r error=TRUE}

filter(data_crane, individual_local_identifier=='L6037')

```

#### b) Make a selection of all observations of crane 'L6037' where the variable `eobs_status` isn't missing.

```{r error=TRUE}

filter(data_crane, individual_local_identifier=='L6037', !is.na(eobs_status))

```

#### c) Make a selection of all observations of crane 'L6037' where the variable `eobs_status` isn't missing. Return only the `event_id` and `timestamp` variables. 

```{r error=TRUE}

data_crane_filter <- filter(data_crane, individual_local_identifier=='L6037')
select(data_crane_filter, event_id, timestamp)

```

### Basic exercise II - Compute the magnitude of the magnetic field

```{r error=TRUE}

data_crane_with_mag_magnetic <- mutate(
  data_crane,  
  magnitude = sqrt(mag_magnetic_field_raw_x ^ 2 + 
                     mag_magnetic_field_raw_y ^ 2 + 
                     mag_magnetic_field_raw_y ^ 2)
)

View(data_crane_with_mag_magnetic)
```


### Reading exercise - Pipe operator

The pipe operator `%>%` is used by many tidyverse users. Read [Chapter 18](https://r4ds.had.co.nz/pipes.html)  of [R for Data
Science](http://r4ds.had.co.nz/). Make a selection columns and rows of the data and make use of the pipe syntax.

### Optional exercise (+) - Exclude variables

```{r error=TRUE}

select(data_crane, -starts_with('location'))

```

### Optional exercise (++) - Summarise results

```{r error=TRUE}

summarise(
  data_crane, 
  min_latitude = max(location_lat, na.rm=T),
  first_observation = min(timestamp),
  magnitude_acceleration = mean(sqrt(acceleration_raw_x^2+acceleration_raw_y^2+acceleration_raw_x^2), na.rm=T)
)
```

### Optional exercise (+++) - Join datasets

```{r error=TRUE}

data_crane_with_measures <- left_join(data_crane, data_crane_additional, by=`event_id`)

```
