---
title: "Modern R with tidyverse [Solutions]"
author: "Jonathan de Bruin, Barbara Vreede"
output:
  pdf_document:
    toc: true
    toc_depth: 4
    latex_engine: xelatex
  html_document:
    toc: true
---


*This document is part of the workshop **Introduction to R & Data** by Utrecht University RDM Support. *


```{r} 
library(tidyverse) 
```

```{r include=F}
# The following code is used to allow errors in code blocks.
knitr::opts_chunk$set(error = T)
```

# 1. Read and save data

## Basic exercise I - Read data into R

### a) Read the data `HUJI_Crane_Israel_GPRS.csv` into R

```{r}
data_crane <- read_csv('data/HUJI_Crane_Israel_GPRS.csv')
head(data_crane)
```

### b) Load `readxl` (to read Excel files)

```{r}
library(readxl)
```

### c) Read the additional observations `crane_additional_observations.xlsx` into R.

```{r}
data_crane_additional <- read_excel('data/crane_additional_observations.xlsx')
head(data_crane_additional)
```

## Basic exercise II - Dataset properties 

```{r}
glimpse(data_crane)
glimpse(data_crane_additional)
```

## Optional exercise (+) - Save data to a CSV file with delimiter `;`. 

```{r}
# create a directory for the output file
if (!dir.exists('tmp')){
  dir.create("tmp")
}

write_delim(data_crane, 'tmp/data_crane_csv_file.csv', delim = ';')
```

## Optional exercise (++) - Write tibble `data_crane` to an Excel file. 

```{r}
install.packages("writexl")
library(writexl)

# create a directory
if (!dir.exists('tmp')){
  dir.create("tmp")
}

write_xlsx(data_crane, "tmp/data_crane.xlsx")

# or, if you want to write to multiple sheets
write_xlsx(list(data_crane = data_crane,
                additional_measurements = data_crane_additional),
           "tmp/data_crane_all.xlsx")
```

## Optional exercise (++) - Read and write SPSS, SAS, and STATA data files

```{r}
library(haven) # to read and write SPSS, STATA and SAS files
```

```{r}
# create a directory
if (!dir.exists('tmp')){
  dir.create("tmp")
}

# read and write SPSS file
write_sav(data_crane, file.path("tmp", "crane_spss.sav"))
read_sav(file.path("tmp", "crane_spss.sav"))

# The SAS file will not write because a column name is too long
# So we perform a quick fix to generate shorter but unique column names
data_crane_sas <- data_crane
names(data_crane_sas) <- str_c(
  str_sub(names(data_crane_sas),1,10), 
  str_sub(names(data_crane_sas),-5),
  sep="_")

# read and write SAS file
write_sas(data_crane_sas, file.path("tmp", "crane_sas.sas7bdat"))
read_sas(file.path("tmp", "crane_sas.sas7bdat"))

# read and write STATA file
write_dta(data_crane, file.path("tmp", "crane_stata.dta"))
read_dta(file.path("tmp", "crane_stata.dta"))
```

## Optional exercise (+++) - Parse datetime columns

```{r}
glimpse(data_crane)
```

```{r}
read_csv('data/HUJI_Crane_Israel_GPRS.csv',
  col_types = cols(
    . = col_guess(),
    eobs_start_timestamp = col_date(format="%Y,%d+%b")
  ))
```


# 2. Data visualisation

## Basic exercise I - Quick plots of the `data_crane`.

### a) Single column plots

```{r}
# for example...
qplot(timestamp, data=data_crane)
qplot(mag_magnetic_field_raw_x, data=data_crane)
qplot(mag_magnetic_field_raw_y, data=data_crane)
qplot(tag_voltage, data=data_crane)
qplot(eobs_status, data=data_crane)
```

### b) Two column plots

```{r}
# for example...
qplot(heading, acceleration_raw_x, data=data_crane)
qplot(acceleration_raw_x, acceleration_raw_y, data=data_crane)
qplot(acceleration_raw_x, acceleration_raw_z, data=data_crane)
qplot(location_long, location_lat, data=data_crane)
```

## Basic exercise II - Using ggplot for plotting

```{r}
p <- ggplot(data_crane, aes(x = heading, y = ground_speed))

p + geom_count()
p + geom_line()
p + geom_point() + geom_density_2d()
```

## Optional exercise (+) - Statistical layers for graphs.

```{r}
p + geom_point() + geom_smooth()
```

## Optional exercise (+) - Scale axes

```{r}
ggplot(data_crane, aes(acceleration_raw_x, acceleration_raw_y)) + 
  geom_point() +
  scale_x_continuous(limits = c(-1000, 1000)) +
  scale_y_continuous(limits = c(-1000, 1000))
```

## Optional exercise (++) - Plot the crane positions on a map

### a) Install the package `maps`

```{r}
install.packages('maps')
library(maps)
```

### b) Plot the crane data on a map.

```{r}
world_map_polygon <- map_data("world2")

ggplot(data_crane) +
  geom_map(data = world_map_polygon, 
           map= world_map_polygon, 
           aes(x=long, y = lat, map_id = region)) + 
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(25, 60)) + 
  geom_point(data = data_crane, 
             aes(x = location_long, y = location_lat))
```

### c) Use an individual identifier to colour the different cranes.

```{r}
world_map_polygon <- map_data("world2")

ggplot(data_crane) +
  geom_map(data = world_map_polygon, 
           map= world_map_polygon, 
           aes(long, lat, map_id = region)) + 
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(25, 60)) + 
  geom_point(data = data_crane, 
             aes(location_long, 
                 location_lat, 
                 colour=individual_local_identifier))
```

## Optional exercise (+++) - Create facets. 

```{r fig.height=15, fig.width=6}
ggplot(data_crane, aes(mag_magnetic_field_raw_x, mag_magnetic_field_raw_y)) +
  geom_point() + 
  facet_wrap(~individual_local_identifier)
```

# 3. Data transformation

## Basic exercise I - Subset data

### a) Filter data from one individual crane

```{r}
filter(data_crane, individual_local_identifier=='L6037')
```

### b) Filter data with complete GPS information

```{r}
data_crane_filtered <- filter(data_crane, 
                              individual_local_identifier=='L6037', 
                              !is.na(eobs_status))

# count the number of rows
nrow(data_crane_filtered)
```

### c) Select specific columns from your data

```{r}
select(data_crane, starts_with("eobs_"))
```

### d) Combine `filter` and `select` using the pipe operator

```{r}
filter(data_crane, individual_local_identifier=='L6037', !is.na(eobs_status)) %>%
  select(starts_with("eobs_"))
```

## Basic exercise II - Compute the magnitude of the magnetic field

```{r}
data_crane_magnetic <- mutate(data_crane, 
                              magnetic_magnitude = sqrt(mag_magnetic_field_raw_x ^ 2 +
                                                          mag_magnetic_field_raw_y ^ 2 +
                                                          mag_magnetic_field_raw_z ^ 2))

# take a look at the new variable
qplot(magnetic_magnitude, data=data_crane_magnetic)
```

## Optional exercise (+) - Summarise results

```{r}
summarise(data_crane,
          min_latitude = min(location_lat, na.rm=T),
          max_latitude = max(location_lat, na.rm=T),
          first_observation = min(timestamp),
          last_observation = max(timestamp),
          magnitude_acceleration = mean(sqrt(acceleration_raw_x ^ 2 +
                                               acceleration_raw_y ^ 2 +
                                               acceleration_raw_z ^ 2), na.rm = T))
```

## Optional exercise (++) - Join datasets

```{r}
data_crane_with_measures <- full_join(data_crane, data_crane_additional, by='event_id')
```

## Optional exercise (+++) - gather all acceleration data

```{r}
data_crane_long <- data_crane %>% 
  filter(!is.na(acceleration_raw_x) & 
           !is.na(acceleration_raw_y) & 
           !is.na(acceleration_raw_z)) %>% 
  gather(acceleration_direction, # column name for headers
         acceleration_speed, # column name for values
         acceleration_raw_x, acceleration_raw_y, acceleration_raw_z # columns to gather
         )

head(data_crane_long)
```