---
title: "Modern R with tidyverse"
author: "[Insert your name]"
output: 
  html_document:
    toc: true
---

*This document is part of the workshop **Introduction to R & Data** by Utrecht University RDM Support. *

# Introduction

In this document, we explore Crane migration, through the GPS data from 39 tagged individuals during the fall migration of 2017. The data was kindly provided for this course by Sasha Pekarsky at the Department of Ecology, Evolution, and Behavior at HUJI. The dataset has been modified by the course instructors.


## Exercises

This part of the workshop (part II) makes use of exercises to find answers to the research questions. The exercises are divided in two groups:

- **basic exercises** The solution of these exercises are part of the rest of this workshop. The code for these exercises needs to work correctly. Ask questions if you are completely stuck.

- **reading exercises** These are reading execises. The reading exercises make use of the book [R for Data Science](http://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham.

- **optional exercises** These questions might be relevant for you. Please pick a few of these exercises to enhance your skills. It doesn't matter if you do not have time to complete them all. The (+) exercises are beginner exercises. (++) Exercises are of a moderate level. The (+++) exercises are really challenging.

## Technical requirements

This project depends on the `tidyverse` package. Load tidyverse into your work environment. 

```{r} 
library(tidyverse) 
```

```{r include=F}
# The following code is used to allow errors in code blocks.
knitr::opts_chunk$set(error = T)
```

# 1. Read and save data

### Basic exercise I - Read data into R

In this exercise, we read the dataset `HUJI_Crane_Israel_GPRS.csv` into the 
memory of the computer. You can find the dataset in the `data` folder of your
project.

#### a) Find the delimiter of the dataset `HUJI_Crane_Israel_GPRS.csv`

The delimiter is the character that separates the values in a CSV file. There
are various ways to find out how the datasets is delimited. Discuss your approach with a colleague or assistant.

#### b) Read the data `HUJI_Crane_Israel_GPRS.csv` into R

Complete the code in the code cell below to read the data into
`data_crane`.

*Tip: The file location (argument `file`) should be 'data/HUJI_Crane_Israel_GPRS.csv'* 

```{r}

data_crane <- read_### enter your code here ###

```

#### c) Load `readxl` (to read Excel files)

The package `readxl` is a tidyverse package to read Excel files. The package
doesn't load by default when you call `library("tidyverse")`. Load the package with the function `library()`. Also read the help function of the `read_excel` function.

```{r}

### enter your code here ###


# open the help function
?read_excel

```

#### c) Read the additional observations `crane_additional_observations.xlsx`.

Read the additional observations in data file `crane_additional_observations.xlsx` (Excel file!). Complete the code in the
code cell below.

```{r}

data_crane_additional <- ### enter your code here ###


```


### Basic exercise II - Dataset properties 

The structure of the dataset (i.e. number of observations, columns, data types) is very important for your initial data exploration. Tidyverse has the function `glimpse()` to output the structure of a `data.frame` or `tibble`. 

Output the structure of both datasets in the cell below. What does `<chr>`, `<dbl>` and `<int>` mean? Are these values what you expect?

```{r}

### enter your code here ###

```

### Reading exercise - `readr` versus base R

You might wonder why we’re not using base R functions for (flat) file reading like `read.csv()`, `read.csv2()` and `read.delimiter()`. These function are available by default in base R and don't require additional packages like `readr`. 

Read [Chapter 11.2.1](http://r4ds.had.co.nz/data-import.html#compared-to-base-r)  of [R for Data
Science](http://r4ds.had.co.nz/) and think about the topics like reproducibility, shareablitiy and performance. Discuss your thoughts with another workshop participant.


### Optional exercise (+) - Save data to a CSV file with delimiter `;`. 

Save the tibble `data_crane` to a `.csv` file. Use `;` as delimiter. 

*Tip: check the function `write_delim`.*
*Tip: Use a different file name, otherwise you overwrite the original data*

```{r}

### enter your code here ###

```

### Optional exercise (++) - Read SPSS, SAS, Excel and STATA data files

(Statistical) software like SPSS, SAS, Excel and STATA are popular programs in the scientific world. These programs do have their own file formats. 

By default, it is not possible to read and write SPSS, SAS and STATA data files with base R. Tidyverse offers solutions to make this possible. The package `haven` can be used to do this (not loaded by default). Load the package with the following line of code:

```{r}
library(haven) # to read and write SPSS, STATA and SAS files
```

#### a) Write tibble `data_crane` to SPSS, SAS, STATA data files. 

Create a code block and do at least one of the following exercises:

SPSS:

- Write the `data_crane` to a SPSS data file (extension `.sav`) with the function `write_sav`.
- Read the saved SPSS data file with the function `read_sav`.

SAS:

- Write the `data_crane` to a SAS data file (extension `.sas7bdat`) with the function `write_sas`.
- Read the saved SAS data file with the function `read_sas`.

STATA:

- Write the `data_crane` to a STATA data file (extension `.dta`) with the function `write_dta`.
- Read the saved STATA data file with the function `read_dta`.


#### b) Write tibble `data_crane` to an Excel file. 

Create a code block and write the tibble `data_crane` to an Excel file (extension `.xlsx`).

Is there a tidyverse function to write Excel files? If not, read on the internet about writing Excel files with R.


### Optional exercise (+++) - Parse datetime columns

Date and time variables can be very challenging to work with in programming languages (including R). In the `data_crane` dataset, there is a variable `eobs_start_timestamp`. This column isn't recognized as a date variable. 

```{r}
glimpse(data_crane)
```

The following code shows how you can parse columns with a specific data type. Insert the correct format for the date column.

```{r}

read_csv('data/HUJI_Crane_Israel_GPRS.csv',
  col_types = cols(
    . = col_guess(),
    eobs_start_timestamp = col_date(format="### insert correct format here ###")
  ))
  
```

# 2. Data visualisation

Visualizing data is important get a feeling for the content of a data set. For the crane dataset, visualisation is used to explore the values of the columns. Base R has strong plotting functions. To get even better flexibility in plotting, we use the ggplot2 functions `qplot` and `ggplot`.

### Basic exercise I - Quick plots of the `data_crane`.

#### a) Single column plots

Make a quick plot of at least 5 columns in the `data_crane` dataset. One example has been provided already.
```{r}

qplot(timestamp, data=data_crane)

### enter your code here ###

```
#### b) Two column plots

Make a few quick plots of combinations of columns. One example has been provided already.

```{r}

qplot(heading, acceleration_raw_x, data=data_crane)

### enter your code here (at least 3 examples) ###


```


### Basic exercise II - Using ggplot for plotting

Complete the following code with different geometric layers. Examples are `geom_point`, `geom_line`, `geom_count` and `geom_density_2d`. 

*Tip: write `geom_` in the search field of the help tab in RStudio to find more geometric layers.

```{r}

ggplot(data_crane, aes(heading, ground_speed)) + 
  ### enter your code here ###

```

### Reading exercise - Statistical layers for graphs.

Statistical layers reveal a strong power of ggplot. The following graph illustrates this:

```{r echo=FALSE, message=FALSE, width=4, height=3}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() + geom_smooth()
```

Read [Chapter 3.6](https://r4ds.had.co.nz/data-visualisation.html#geometric-objects)  of [R for Data
Science](http://r4ds.had.co.nz/) and experiment with the function `geom_smooth()` and `stat_smooth()`. What is the difference between both functions? 


### Optional exercise (+) - Scale axes

Scaling of the axes with a ggplot is easy. Take a look at the *data-visualization* cheatsheet. In the bottom right corner of the cheatsheet, you will find the code needed to scale the axes. 

Zoom in on the acceleration between -1000 and 1000 for both the x-axis and y-axis. 

```{r}
ggplot(data_crane, aes(acceleration_raw_x, acceleration_raw_y)) + 
  geom_point()
```



### Optional exercise (++) - Plot the crane positions on a map

In this exercise, we plot the crane data on a map.

#### a) Install package the `maps`

Install the package `maps` and load the library in the cell below.

```{r}

### enter your code here ###

```

#### b) Plot the crane data on a map.

Add a layer with the location of the observations.

*tip: can you use the `geom_point()` layer for this?*


```{r}
world_map_polygon <- map_data("world2")

ggplot(data_crane) +
  geom_map(data = world_map_polygon, map= world_map_polygon, aes(x=long, y = lat, map_id = region)) + 
  scale_x_continuous(limits = c(0, 60)) +
  scale_y_continuous(limits = c(25, 60))
  ### add your geometric layer here ###

```


#### c) Use an individual identifier to colour the different cranes.

The variable `individual_local_identifier` identifies the cranes. Use this variable to color the points.

```{r}


```

### Optional exercise (+++) - Create facets. 

Search on the internet for information about facets in ggplot. This unique feature is a must-have for data exploration. Create a graph (it doesn't matter what kind of graph) and use facets to display all cranes (use the variable `individual_local_identifier`). 

```{r fig.height=15, fig.width=6}

### enter your code here ###

```


# 3. Data transformation

The crane dataset is clean and well-structured. Nevertheless, you might be interested in doing additional transformations and selections. In this section, we transform and enrich the crane dataset. 

### Basic exercise I - Subset data (single crane)

#### a) Make a selection of all observations of crane 'L6037'.

Make a selection of observations of crane 'L6037' in column `individual_local_identifier`. Make use of the tidyverse (dplyr) function `filter()`. Have a look at the cheat sheet if you are stuck.

```{r}

### enter your code here ###


```

#### b) Make a selection of all observations of crane 'L6037' where the variable `eobs_status` isn't missing.

How many records satisfy this condition?

*Tip: Use the function `is.na()` preceded by the `!`. Check the help page for both `is.na` and `!`.*

```{r}

### enter your code here ###

```

#### c) Make a selection of all observations of crane 'L6037' where the variable `eobs_status` isn't missing. Return only the `event_id` and `timestamp` variables. 

Make use of tidyverse functions only.

```{r}

### enter your code here ###

```

### Basic exercise II - Compute the magnitude of the magnetic field

The function `mutate()` is a very powerful function in tidyverse (the `dplyr` package) to mutate and create new column variables in a data frame or tibble. 

Create a tibble with a new variable called `mag_magnetic`, which is the magnitude of the magnetic field. You can calculate the magnetic field by taking the square root of `mag_magnetic_field_raw_x ^ 2 + mag_magnetic_field_raw_y ^ 2 + mag_magnetic_field_raw_y ^ 2`  

```{r}

data_crane_with_mag_magnetic <- mutate(data_crane,  ### enter your code here ### )

head(data_crane_with_mag_magnetic)
```

Try to use the formatting discussed during the presentation, so that your code is as readable as possible.

### Reading exercise - Pipe operator

The pipe operator `%>%` is used by many tidyverse users. Read [Chapter 18](https://r4ds.had.co.nz/pipes.html)  of [R for Data
Science](http://r4ds.had.co.nz/). Make a selection columns and rows of the data and make use of the pipe syntax.

### Optional exercise (+) - Exclude variables

Create a tibble of the crane dataset without the latitude and longitude. 

Use tidyverse and a maximum of 60 characters.

```{r}

### enter your code here ###

```

### Optional exercise (++) - Summarise results

The function `summarise()` can be used to summarise variables in tibbles. Compute the following properties: 

- the maximum latitude (the highest point on the globe).
- the first observation
- the magnitude of the acceleration (the square root of x^2 + y^2 + z^2)

*Tip: Use the cheatsheet or the book R4DS*

```{r}

summarise(
  data_crane, 
  min_latitude = ### enter your code here ###
  first_observation = ### enter your code here ###
  magnitude_acceleration = ### enter your code here ###
)
```

### Optional exercise (+++) - Join datasets

Quickly read [13.4.1](https://r4ds.had.co.nz/relational-data.html#understanding-joins)-[13.4.3](https://r4ds.had.co.nz/relational-data.html#outer-join) from R for Data Science. Focus on the images.

In this exercise, append crane dataset with the additional measures. What kind of join is this? *Hint: Check the data transformation cheatsheet, section 'Combine Tables'.*

The join key for our datasets is `event_id`. 

```{r}

# data_crane_with_measures <- ..._join()

```

Multiple options give the same results. Explain why!
