---
title: "Exercises in BaseR"
output:
  html_document
---
```{r include=F}
# The following code is used to allow errors in code blocks.
knitr::opts_chunk$set(error = T)
```

Welcome! You are looking at an R markdown file containing Base R exercises. 
R markdown files contain both text and R code. This allows us to provide textual instructions as well as code blocks where you can type and execute the answer. 
In regular R scripts (.R files), text has to be preceded by a # so that R will 
not run that line. This is not so handy here because there is so much text!

#### Exercise 0: running R code

Try to run this code, by clicking the green 'Play' button.

```{r}
x <- 6

apple <- "apple"
```


### Exercise 1: R syntax & data types

**Write your code into the code chunk (the grey area) and press the green 'Play' button to execute.**

1. Do the following calculation in R: one plus five, divided by nine.

2. Assign the result of the calculation to a variable.

3. Test if the result of your calculation is larger than 1.

4. Round off the result to 1 decimal. Tip: use the Maths Functions section of the Base R cheat sheet!


```{r}


```


### Exercise 2: Create vectors

1. _Meet Ann, Bob, Chloe, and Dan._ 
Create a character vector called "name" and add these names to the vector using the `c()` function.

2. _How old are Ann, Bob, Chloe, and Dan?_ 
Create a numeric vector called "age" and add their ages to the vector. You can decide their ages.

3. Use the `class()` function to check the data type of the `name` vector. Do the same for the `age` vector.

4. _What is their average age?_ 
Use a function in R to calculate this. 
**Tip:** use the Maths Functions section of the Base R cheat sheet!

```{r}


```


### Exercise 3: Combining data

1. Create a vector called "country" containing four countries of your choice.

2. Create a data frame called "df" combining name, age, and country.

3. Create a list called "mylist" with the 3 vectors and 1 dataframe you have just created. 

```{r}


```

### Exercise 4: Selecting vector & list elements

1. Return only the first number in your vector age.

2. Return the 2nd and 4th name in your vector name.

3. Return only ages under 30 from your vector age.

```{r}

```


### Exercise 5: Selecting from a dataframe

Before you start, please run this code:
```{r}
rm(name,age,country)
```

1. From your dataframe df, return complete rows for everyone living in a country of your choice.

2. Return only the names of everyone in your data frame df under 40. 

3. Return the columns name and age together.
	
```{r}

```


### Exercise 6: Predict the answer

Predict the results before you run the code. Does the real answer make sense to you?

```{r}
5==5
``` 

```{r}
5==NA
```

```{r}
NA==NA
```

```{r}
is.na(NA)
```


### Exercise 7: If statement

Make an if statement that tests if a number is larger than 18. 
If the number is larger, the variable age_category should be assigned the value "adult". If not, the variable age_category should be assigned the vaue "minor".

```{r}
number <- 8

if(){
  
}

print(age_category)
```

### Exercise 8: Function

Turn the if statement from the last exercise into a function. Let the user provide the value for number, and return the age_category.

After you've created your function, test it with different inputs!

```{r}

test_age <- function(){
  
}


# Test the function
test_age(<your-input>)
```

### Exercise 9: For-loop

Go over the age column in your dataframe df, and for each age: print() the age category using the test_age function from the previous exercise.

```{r}
for( in ){
  
}

```

Bonus question!

Create a different for-loop. This for-loop runs the test_age function for each age in your dataframe df,
and adds the age categories as a new column to your dataframe df.

```{r}
# We first set an index that will be increased every time the for-loop runs
i <- 1

for(){
  # Add the age category to a new column in df
  ???
    
  # Increase the index with 1 after running the code in this for-loop iteration
  i <- i + 1
}

```
