---
title: "Data Science with the Tidyverse"
output:
  html_document:
    toc: true
---
<!-- This code block sets some defaults for the code chunks, and can safely be ignored! -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


_This document is part of the workshop **Introduction to R & Data** by Utrecht University RDM Support._

# Introduction

In the following exercises, we will work with a dataset on penguins, collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER, a member of the Long Term Ecological Research Network.

The dataset was processed for R by Allison Horst, and is available for use under a CC0 license. For more information on this awesome dataset, check out [the palmerpenguins repository](https://allisonhorst.github.io/palmerpenguins/).

## Exercise 0

Load the `tidyverse` package.

```{r} 
library(tidyverse) 
```

# Chapter 1: Importing data

## Exercise 1 

In this exercise, we read the dataset `penguins.tsv` into R. 

You will need to know the **location** of your data (hint: check the `data` folder),
and the **function** to use to read data into R.

Complete the code in the chunk below, replacing the ??? with the with the correct code:

```{r}
penguins <- read_tsv("???")
```

## Exercise 2

`read_tsv` and similar functions like `read_csv()` are useful functions for most tabular data files, but they are unable to read Excel files. For this, we need an additional package called `readxl`.

`readxl` was installed when you installed `tidyverse`, but it does not load by default when you call `library(tidyverse)`. 

So, we first need to load this package with the function `library()`:

```{r}
library(readxl)
```

Use the function `read_excel()` to read a related data set `penguins_isotopes.xlsx` (an Excel file!), also located in your `data` folder, into R.

```{r}
penguins_isotopes <- ???(path = "???")
```

## Exercise 3

Save your Excel data as a _comma separated values_ file, with the function `write_csv()`.

For this function, you need to provide:
- The name of the R object you want to save: `penguins_isotopes`
- The name of the folder and the file name: "data/penguins_isotopes.csv"

```{r}
???(penguins_isotopes,
    # Name of the file, including extension (.csv)
    file = "???")
```

# Chapter 2: Selecting & Filtering Data

We are going to clean up our data by selecting/filtering only the relevant information.

## Exercise 4

For some of the penguins, the sex was not determined. Use the function `filter()` to keep only the rows where the column `Sex` specifies either "FEMALE" or "MALE".

One way to do this (though not the only way) is to filter out the rows where `Sex` is NA. Remember the function `is.na()` and remember that to negate a condition, you use the operator! For example, `!is.na(NA)` returns `FALSE`.

```{r}
penguins_subset <- filter(penguins, ???)
```

## Exercise 5

We are going to select columns with relevant data, for our upcoming analysis.
The research we do is on penguin bill sizes (described in our data with the word "Culmen").

We want to include a few relevant variables so we can analyse them later:
- Individual_ID
- Species
- Sex
- Island
- Culmen_Depth_mm
- Culmen_Length_mm

Use the function `select()` to include the right columns in our data subset. 

**Note** that in order to work with the subset from Exercise 4, we need to use that data frame (`penguin_subset`) as our data input!

```{r}
penguins_subset_2 <- select(???)
```

## Exercise 6

Use the `mutate()` function to create a new column called `culmen_ratio`, which is calculated as `Culmen_Length_mm` divided by `Culmen_Depth_mm`.

Make sure to work with the `penguins_subset_2` dataframe.

```{r}
penguins_subset_3 <- mutate(penguins_subset_2, ???)
```

## Exercise 7

In the next chapter, we will make our dataframe tidy. Before we do that, we will rename the `Culmen_Length_mm` and `Culmen_Depth_mm` columns in order to make subsequent operations easier.

```{r}
penguins_subset_4 <- rename(penguins_subset_3,
                            length = ???,
                            depth = ???)
```

# Exercise 8

Use %>% (the pipe operator) to combine the code you wrote in Exercises 4, 5, 6 and 7. 

Make a workflow that starts with the data `penguins` and subsequently applies your `filter`, `select`, `mutate` and `rename` operations. Make sure to remove references to the original dataframe in subsequent operations, as this is now provided through the %>% operator.

```{r}
penguins_subset_5 <- penguins %>%
  ??? %>%
  ??? %>%
  ??? %>%
```

# Chapter 3: Tidy Data

## Exercise 9

Now that we've renamed our columns, we will transform the dataframe from wide to long format using the function `pivot_longer()`:

For this, you need to specify:
- what columns are going to become rows (under the argument `cols`)
- where the original column names will go using `names_to`
- where the values of the original column will go using `values_to`

```{r}
penguins_long <- penguins_subset_5 %>%
  pivot_longer(cols = c(???, ???),
               names_to = "culmen_element",
               values_to = "measurement")
```

### Exercise 10 

Use the functions `group_by()` and `summarize()` in conjunction to calculate the mean and standard deviation of all measurements, grouped by species and the type of measurement.

- Within `group_by()`, specify `Species` and `culmen_element`

- Within `summarize()`, use the functions `mean()` and `sd()` to calculate the average and standard deviation of the column `measurement`.

Make sure to feed the right data frame into the workflow: the column `culmen_element` is one that was created in the previous exercise!

**Note:** a warning message may appear that says: "`summarise()` regrouping output by 'Species' (override with `.groups` argument)". This is expected behavior, don't worry!

```{r}
penguins_summary <- penguins_long %>%
  group_by(???) %>%
  summarize(avg = ???,
            sd = ???)
```

## Exercise 11 

Use the function `full_join()` to merge the `penguins_summary` and `penguins_long` data frames, so that each row in the long data frame will have additional columns with the mean and standard deviation for its group.

```{r}
penguins_join <- full_join(???)
```

# Chapter 4: Data Visualization

## Exercise 12

Using the ggplot2 package, let's plot the culmen length against their flipper length from the `penguins` dataframe.
Culmen_Length_mm and Flipper_Length_mm are both continuous variables, and therefore we are now choosing a scatterplot.

In the code chunk below:

- Put Culmen_Length_mm on the x-axis, and Flipper_Length_mm on the y-axis
- Color the points according to Species
- Add scatter points using the geom `geom_point()`
- Add x- and y-labels and a title with `labs()`
- Add a 'theme' layer to your plot. There are many default themes available, check them out here: https://ggplot2.tidyverse.org/reference/ggtheme.html.
- Bonus 1: Use `facet_wrap()` to make subplots per Island. The `~` needs to be used before the variable that you wish to subplot.
- Bonus 2: Use `geom_smooth()` to add regression lines to the scatter plots with the method "lm" (linear model)

```{r}
ggplot(penguins, aes(x = ???, y = ???)) +
  ???
```


## Exercise 13

Create another type of data visualization with any of the following dataframes that we've created: `penguins`, `penguins_long`, `penguins_summary` or `penguins_join`. 

Tip: Use the ggplot2 cheatsheet, or find a nice data visualization (and code examples) at https://www.data-to-viz.com/.

```{r}
# Add your code to create your plot here!

```
